Staged Abstract Interpreters

### 2019-08-12 paper/10_intro.tex
revision

\paragraph{Contributions} Briefly, the contribution of the paper is as follows:
\begin{itemize}[leftmargin=2em]
  \item To our best knowledge, we present the first one to use
    multi-stage programming to specialize a general abstract interpreter,
    and our approach does not touch the soundness of analyses.
  \item Intellectually, our framework naturally extends the first
    Futamura projection to abstract interpreters, showing a
    well-grounded approach to optimize static analyses via
    metaprogramming.
  \item Pragmatically, we show that staging abstract interpreters is
    useful to improve performance and scalability of analyses by case
    studies and empirical evaluation.
\end{itemize}

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our core target language and reviewing
    monads in Scala, and then presenting the generic interpreter (Section~\ref{prelim}).
    After which, we review its instantiations of concrete interpretation
    (Section~\ref{unstaged_conc}) and staged concrete interpretation
    (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same framework by
    replacing the environment, store and values (Section~\ref{unstaged_abs}).
    After, we show the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
  \item We conduct three case studies (Section~\ref{cases_study}), which demonstrate that
    our approach requires less engineering efforts, is applicable to various analysis,
    and enables modular analysis.
  \item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

>>>>

\paragraph{Contributions} Briefly, the contribution of the paper is as follows:
\begin{itemize}[leftmargin=2em]
  \item To our best knowledge, we present the first one to use
    multi-stage programming to specialize a general abstract interpreter,
    and our approach does not touch the soundness of analyses.
  \item Intellectually, our framework naturally extends the first
    Futamura projection to abstract interpreters, showing a
    well-grounded approach to optimize static analyses via
    metaprogramming.
  \item Practically, we show that staging an abstract interpreter is
    useful to improve performance and scalability of analyses by case
    studies and an empirical evaluation.
\end{itemize}

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter (Section~\ref{prelim}).
    After which, we review its instantiations of concrete interpretation
    (Section~\ref{unstaged_conc}) and staged concrete interpretation
    (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same framework by
    replacing the environment, store and values to their abstract counterparts (Section~\ref{unstaged_abs}).
    After, we show the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
  \item We conduct three case studies (Section~\ref{cases_study}), which demonstrate that
    our approach requires less engineering efforts, is applicable to various analysis,
    and enables compiling analysis modularly.
  \item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

### 2019-08-12 paper/20_prelim.tex
intro
revision

In this section, we first describe the abstract syntax of the language for our interpreter, 
then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative/direct style and other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

\iffalse
Then we
will proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).
\fi

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use algebraic data-types to represent programs. This design choice allows us to
easily use different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

>>>>

In this section, we first describe the abstract syntax of the language for our interpreter, 
then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative or direct style in other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use data-types to represent programs. This design choice allows us
to easily use different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language constructs and reuse
existing interpretations \todo{cite Bruno?}.

### 2019-08-12 paper/50_stageabs.tex
revision

\subsection{Discussion}

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we discuss several decisions we made to
achieve this and examine some alternatives.

\paragraph{What does the generated code look like?} In the generated code, all the
primitive operations (such as @eval@, @fix@, @ap_clo@, etc.) and monadic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representations are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

@@

\paragraph{Correctness and Soundness}

>>>>

\section{Discussion}

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we review and summarize our recipe
to achieve the staged abstract interpreter, discuss the correctness issue
and different design choices.

\subsection{Summarizing the Approach}

\todo{TODO}

\paragraph{What has been eliminated?} In the generated code, all the
primitive operations (such as @eval@, @fix@, @ap_clo@, etc.) and monadic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representations are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

\todo{may be show an small example?}

\subsection{Correctness}

\todo{TODO}

\subsection{Design Choices}

@@

### 2019-08-12 paper/80_related.tex
update
revision

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The AAM approach is closely related to control-flow analysis
for higher-order languages \cite{Midtgaard:2012:CAF:2187671.2187672}.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
Similar to definitional abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis.

\paragraph{Meta-Programming} In this paper, we use the Lightweight Modular
Staging framework \cite{DBLP:conf/gpce/RompfO10}, which relies on types as stage
annotations. Other notable implementations of MSP exist in ML family, e.g.,
MetaML \cite{DBLP:conf/pepm/TahaS97} and MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
The idea of staging an abstract interpreter we presented in this paper is also
applicable to other MSP systems. Partial evaluation as an automatic technique is
studied comprehensively \cite{10.1007/3-540-61580-6_11,
DBLP:books/daglib/0072559}. Futamura projections reveals close relations between
interpreters and compilers, which was originally proposed by
\citeauthor{futamura1971partial} in 1970s\cite{futamura1971partial}, and later
reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} shows how to collapse a tower of
concrete interpreters by MSP -- it would be also interesting to explore multiple
layers of abstract interpreters.
\citet{10.1007/11561347_18} discussed combining
multi-stage programming with functors and monads.
Similar to the idea in this paper, specializing monadic interpreters was
explored by \citet{DBLP:conf/dsl/SheardBP99, danvy1991compiling}.

Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.

>>>>

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The AAM approach is closely related to control-flow analysis
for higher-order languages \cite{Midtgaard:2012:CAF:2187671.2187672}.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
Similar to definitional abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis, \todo{but intentional properties
  are different, compare with summary}

\paragraph{Two-level Semantics}

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation as an automatic technique for program specialization
was studied comprehensively by \citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.
In this paper, we use mutli-stage programming as an approach to
program specialization. The Lightweight Modular Staging framework \cite{DBLP:conf/gpce/RompfO10}
we used in the paper relies on type-level stage annotations.
Other notable implementations of MSP exist in ML family, e.g.,
MetaML \cite{DBLP:conf/pepm/TahaS97} and MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
The idea of staging an abstract interpreter we presented in this paper is also
applicable to other MSP systems.
Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.

As an source of inspiration of this paper, Futamura projections reveals a
hierarchy and close relation between interpreters and compilers, which was
originally proposed by \citeauthor{futamura1971partial} in 1970s\cite{futamura1971partial},
and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} consider a tower of concrete
interpreters and how to collapse them by using MSP -- it would be
interesting to explore this idea for multiple layers of abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed combining
multi-stage programming with functors and monads. But they didn't \todo{monad transfomers?}
Similar to the idea in this paper, specializing monadic interpreters was
explored by \citet{DBLP:conf/dsl/SheardBP99, danvy1991compiling}. \todo{what's the difference?}

\paragraph{Control-flow Analysis}
